Modeling memory and taxis to analyze migrations

### 2021-07-10 text/ms.Rmd

*Highly sketchy and incomplete.  Some odd thoughts here and there.*

We found that the relatively straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration. In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location. The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether as trends or inter-annual stochasticity.  Importantly, we were not interested here in a process that do *not* correspond to the "green wave surfing" mechanism, as there is very little trend to surf, but consider only a necessarily strategic, long-scaled displacements and the way those displacements are modified by experience.

The ability to migrate depended on the nature of the resource dynamics.  In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.  Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.  Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.

It is worth considering the assumptions of our model with respect to a biological population.  The population is explicitly social and collective, and the memory is a memory that belongs to the entire population.  The importance of the spatial scale of the collective was an unexpected and very strong result, suggesting that for a population to adequately learn socially, it must have a large enough exploratory radius to find new information to adapt to, while being small enough that the entire population can then learn from this.

>>>>

Animals navigate complex, dynamic, patchy environments.  When there is a strongly localized and seasonal component to the resource dynamics, typical strategies of spatial distribution by straightforward resource-following taxis necessarily fails to maximally exploit the dynamics.  It is in these cases - quite common in the natural world - that seasonal migration becomes a viable, even necessary, strategy.  However, when those resources start shifting in space and time - as is occurring at an accelerated pace with recent global climate change - the migration phenology itself must exhibit some plasticity.  It is our conjecture that this plasticity is greatly facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years.  To our knowledge, this study is the first attempt to focus in and propose a relatively simple, memory-driven mechanism by which a population can respond to long-term changes in resource dynamics.

We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration. In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location. The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity.  We note that our model did not include any selection, inheritance or birth or death processes, i.e. there was evolutionary optimization of parameters to maximize a fitness, as in several somewhat similar models that explicitly explore the evolution of migration [@Guttal2010;@Shaw2013;@Anderson2013].  Rather, we postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior. The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.

It is worth considering the bioloogical interpretations of the various parameters. The diffusion parameter ($\epsilon$) captures short time-scaled randomness and exploration of a population and the foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.  These are the most common ingredients in diffusion-advection models of animal movement and direct parallels can be drawn between these and, respectively, individual random movement models [@Gurarie2011] and resource selection models, which are the fundamental ingredients of empirical analysis of animal movements. The sociality parameters, which turn out to be exceedingly important for the resiliency of this process, capture two very important but lesser-studied properties of animal movements.  The strength of sociality ($\beta$) essentially quantified the importance of the group.  Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion; and it appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology. The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, and was, somewhat unexpectedly, among the most important parameters.  In particular, a small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak. This result points to an important fundamental dynamic: that in order to adapt to any shift in resources, some individuals must experience that resource, and the two sociality-related mechanisms to have that scope is either a relatively large spatial scale or a relatively weak social cohesion.

The ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics.  In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.  Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.  Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.
